Asymmetric relationships and their effects on coexistence

### 2023-09-11 paper/RefereeResponses2.tex

\noindent \emph{Yours,}

\noindent \emph{Dan Reuman and Jasmin Albert} \\
%DONE with cover letter

\noindent \textbf{Comments of referee 1:} \\

\noindent C1) First, to consider introducing a discussion of the robustness to more esoteric definitions of the partial-sharp distributions, and adding this briefly to the main text.  Primarily as the authors already suggest in their response letter, as a signpost for future work.\\

\noindent ***\emph{} \\
%DONE

\noindent C2) The only other minor suggestion is that the authors clarify that in Fig 1 they are simply showing that ATAs do occur in nature.  I was not suggesting that the authors include examples pertinent to their question in Figure 1---as they say, readers may not be expecting this. I don't think that the connection between the ATAs in Fig 1 and the rest of the project are that clear, as I already expressed. But I respect the authors wishes (and their other feedback) that it is worthwhile including these. Just a sentence or so at some point explaining that these are not directly related to the ATAs relevant in the remainder of the paper may avoid confusion. \\

\noindent ***\emph{} \\

@@

\noindent \textbf{Comments of referee 2:} \\

\noindent C3) This is a novel contribution highlighting that asymmetric tail associations in correlations between two variables was a very simple aspect not properly considered before but with huge implications for the maintenance of biodiversity. I congratulate the authors for their work. \\

@@

\noindent ***\emph{Thank you for the very positive feedback!} \\
%DONE

\noindent \textbf{Comments of referee 3:} \\

\noindent C4) I thank the authors for the change and the nice discussion/replies to my questions. I especially like the new figure 2 which greatly helped my understanding of how to remove ATA. \\

\noindent ***\emph{Many thanks for the positive feedback!} \\

@@

\noindent \textbf{Comments of the editor:} \\

\noindent C5) Reviewers are enthusiastic about this paper demonstrating the importance of how variable relationships depending on when they are weak or strong influence competitive exclusion and coexistence. \\

\noindent ***\emph{Thank you for pointing out.} \\

\noindent C6) The legends of figures are rather long and dense. Figures should be self-explanatory. I suggest adding to all of them the sentence at the end: “See Table 1 for used annotations and abbreviations.” \\

\noindent ***\emph{We }

\noindent C7) The legend in Fig. 1 is too long and has variables not present in the graphs but mentioned in other sections of the paper. For example, in line 44-45: “The variables (B1, B2) described in the Introduction…”. I suggest revising if this information is needed at all in this Figure. \\

\noindent ***\emph{Thank you for the very positive feedback!} \\

\noindent C8) Main concepts should be described in the introduction in a way that could be understood for a broad audience. In this sense, I suggest adding a short non- mathematical definition for storage effects even if you go in more detail in the Theory section in lines 59-60 \\

\noindent ***\emph{Thank you for the very positive feedback!} \\

\noindent C9) Page 2, Lines 59-60 “coexistence into the contributions of various mechanisms, with names such as storage effects (Chesson1994). Storage effects refer to….. Storage effects are reviewed conceptually below and defined formally in Theory. \\

\noindent ***\emph{Thank you for the very positive feedback!} \\

\noindent C10)  Page 2, Line 60: Start a new paragraph with the sentence “MCT has been applied to several systems…” \\

\noindent ***\emph{Thank you for the very positive feedback!} \\

### 2023-09-11 paper/Paper.Rnw
responding to editor's comments

%Intro to competition, coexistence, MCT, etc.
Competition and coexistence are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}. The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will 
instead drive the other extinct \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems,
through the niche partitioning effects of environmental variation.
MCT breaks down coexistence into the contributions of various mechanisms, with names such as 
``storage effects'' \citep{Chesson_1994}; storage effects are reviewed conceptually below and 
defined formally in Theory. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Caceres_1997, Descamps_2005, Letten2018}, and corals \citep{Alvarez-Noriega_2020}. 
%Narawani_2013
Although MCT is somewhat mathematically complex, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig1_a-h.pdf}
\caption{Examples motivating our study of the effects of asymmetric tail associations (ATAs) on competition
and coexistence. (a-c) Bivariate random variables with standard normal marginals showing
alternative patterns of association in the left and right parts of the distributions. 
The two variables in (a) are more strongly related to each
other in the left parts of the distributions, termed ``left-tail association;'' those in (b) 
are symmetrically associated; and those in (c) 
are more strongly related to each other in the right parts, termed ``right-tail association''. All three cases have 
the same Pearson correlation, $\rho$, up to sampling variation, so Pearson correlation does not identify these 
very different patterns of association. For (a, c), we used asymmetrically tail associated cases which are perfectly 
correlated in the lower or upper halves of the distributions, though it is also possible to generate asymmetrically 
tail associated noise with imperfect correlations in both halves of the distributions [\citealp{Nelsen2006,Ghosh_2020_AER};
see also (d,e)]. 
The variables $(B_1,B_2)$ described in the Introduction and Methods were obtained by generating noise as 
pictured and then transforming the marginals as specified in Methods.
(d,e) Two examples of ATAs found in nature. Axes are population densities of two plankton species, 
\emph{Ceratium fusus} and \emph{Ceratium furca}, sampled in successive years in two of the 15 locations 
in the North Sea considered by \citep{Ghosh_2020_ecolevol}. ATAs were statistically
significant and differed by location, with associations being primarily in the left tails in 
panel d, and primarily in the right tails in panel e (SI section \ref{SI-SIsect:planktonstats}). 
One can see these asymmetries by noting that the open points in panel d, which are in the left halves of the
distributions, are more tightly clustered around the line $y=x$ than are the closed points,
which are in the right halves of the distributions; and \emph{vice versa} for panel e.
(f-h) Lottery model population simulations using $(B_1,B_2)$ obtained by transforming the 
respectively corresponding variables (a-c; see Introduction).
\label{fig:pedagog1}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/theory_fig_col2.pdf}
\caption[Visual guide to removing asymmetries of association]{Illustration of how asymmetric tail associations are removed while retaining marginal distributions and correlation 
between variables (see Theory). (a) An example original bivariate distribution $(E_i,C_{i \bs i})$, with ATAs present. 
(b) Marginals are now normalized. The variables $E_i$ and $C_{i \bs i}$ are individually transformed to have standard normal 
marginals through a composition of the inverse of the standard normal cumulative distribution function (CDF), 
$\varphi^{-1}$, and 
each variable's CDF, $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively.  
(c) A newly generated bivariate normal distribution, $(e_i,c_{i \bs i})$, with correlation $\ddot{\rho}=\rho$
selected to match that of the original distribution (see Theory). (d) Two new variables, 
$E_i^{\nat}$ and  $C_{i \bs i}^{\nat}$,
which have the same marginals and correlation as the original $E_i$ and $C_{i \bs i}$ variables of (a) but 
which have symmetric association structure. These are the distributions used to compute ATA effects, 
$\Delta_i^{[EC]}$ (Theory). 
These variables are obtained by transforming the marginals from (c), 
$e_i$ and $c_{i \bs i}$, using the inverses of the functions used in (b).
Panel (e) shows the distribution $(E_i^\shp,C_{i \bs i}^\shp)$, defined in Theory
following \cite{Ellner_2016}, for comparison. These are the distributions used to compute storage effects,
$\Delta_i^{(EC)}$ (Theory).
For illustration purposes we started with beta distributed $E_i$ and $C_{i \bs i}$. Contour lines are 
$\log_{10}$ of probability density on all panels. 
\label{fig:pedagog2}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLLN.pdf}
\caption{The contribution of ATAs to species coexistence was sometimes, but not always, 
meaningfully large compared to 
other components of the MCT decomposition for the log-normal fecundities lottery model.
All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel, 
$\delta$ and $\mu_1-\mu_2$ 
are fixed at values specified in the margins, and GWR and components of the decomposition
are plotted against $\sigma$. The contribution of ATAs ($\Delta_i^{[EC]}$) sometimes, but not always, 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the GWR line falls 
between the ATA line and the $x$ axis, indicating that, for those parameter values, ATAs 
make the difference between species
coexistence and exclusion of the weaker competitor. These instances, which are shaded, correspond to
ATA rescue or exclusion (Theory). Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3,5)}, so are not displayed. This plot uses $q_{ij}=1$. An 
analogous plot (SI Fig. \ref{SI-fig:fig3_alt}) uses 
$q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$; though results differed
in detail, main conclusions were the same. \label{fig:fig3}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLB.pdf}
\caption{The contribution of ATAs to species coexistence was often meaningfully large compared to 
other components of the MCT decomposition for the beta fecundities lottery model.
All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel,
$\delta$ is fixed at the value specified in the upper margin, 
and GWR and components of the decomposition
are plotted against the upper bound ratio, $\eta_2/\eta_1$. The contribution of ATAs ($\Delta_i^{[EC]}$) frequently 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the GWR line falls 
between the ATA line and the $x$ axis, indicating that, for those parameter values, ATAs make 
the difference between species
coexistence and exclusion of the weaker competitor. These instances, which are shaded, correspond to
ATA rescue or exclusion (Theory). Standard errors of plotted quantities 
were never greater than \Sexpr{round(se4,5)}, so are not displayed. 
This plot uses $q_{ij}=1$. 
An analogous plot (SI Fig. \ref{SI-fig:fig4_alt}) uses 
$q_{ij}=\eta_i/(\eta_j(1-\delta)+\eta_i \delta)$; 
though results differed in detail, main conclusions were the same.
}\label{fig:fig4}
\end{figure}

@@

\noindent Figure 5 caption. ATA effects and ATA rescue in the diatom system. 
The contribution of ATAs to the GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
other components of the MCT decomposition (a-c). Panels a-c show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
For each panel, 
two of the three parameters defining the temperature oscillation (the amplitude, $a$; period, $P$; 
and mean, $\theta_0$; see Methods) were fixed at the values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, $\theta_0=18^\circ$C), and 
GWR and components of the MCT decomposition
were plotted against the third parameter. The original values of the varying parameters 
are indicated by vertical dotted lines. 
ATA contributions ($\Delta_i^{[EC]}$)
had magnitude comparable to other components, 
indicating that ATAs can contribute meaningfully
to coexistence in this system. In some instances, the GWR line falls 
between the ATA line
and the horizontal axis, indicating that, for those parameter values, ATA rescue occurs. 
Standard errors of plotted quantities were always less then \Sexpr{round(se5,5)}, so are not shown.
Panels d-f show bivariate plots (Methods) for which the third parameter was 
fixed at the values used by \cite{Descamps_2005}. ATAs are again often an important factor helping
determine coexistence or exclusion, and ATA rescue happened for an appreciable fraction of the
explored parameter space.
For the parameters we considered, GWR of the other diatom species in the system, 
\emph{C. pseudostelligera}, was always positive both including and
excluding the effects of ATAs, so GWR for \emph{F. crotonensis}, decomposed here, is a 
sufficient coexistence metric.

>>>>

%Intro to competition, coexistence, MCT, etc.
Competition and coexistence are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}. The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will 
instead drive the other extinct \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems,
through the niche partitioning effects of environmental variation.
MCT breaks down coexistence into the contributions of various mechanisms, with names such as 
``storage effects'' \citep{Chesson_1994}. 
Storage effects refer to ... and are reviewed conceptually below and 
defined formally in Theory.

MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Caceres_1997, Descamps_2005, Letten2018}, and corals \citep{Alvarez-Noriega_2020}. 
%Narawani_2013
Although MCT is somewhat mathematically complex, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig1_a-h.pdf}
\caption{Examples motivating our study of the effects of asymmetric tail associations (ATAs) on competition
and coexistence. (a-c) Bivariate random variables with standard normal marginals showing
alternative patterns of association in the left and right parts of the distributions. 
The two variables in (a) are more strongly related to each
other in the left parts of the distributions, termed ``left-tail association;'' those in (b) 
are symmetrically associated; and those in (c) 
are more strongly related to each other in the right parts, termed ``right-tail association''. All three cases have 
the same Pearson correlation, $\rho$, up to sampling variation, so Pearson correlation does not identify these 
very different patterns of association. For (a, c), we used asymmetrically tail associated cases which are perfectly 
correlated in the lower or upper halves of the distributions, though it is also possible to generate asymmetrically 
tail associated noise with imperfect correlations in both halves of the distributions [\citealp{Nelsen2006,Ghosh_2020_AER};
see also (d,e)]. 
%The variables $(B_1,B_2)$ described in the Introduction and Methods were obtained by generating noise as 
%pictured and then transforming the marginals as specified in Methods.
(d,e) Two examples of ATAs found in nature. Axes are population densities of two plankton species, 
\emph{Ceratium fusus} and \emph{Ceratium furca}, sampled in successive years in two of the 15 locations 
in the North Sea considered by \citep{Ghosh_2020_ecolevol}. ATAs were statistically
significant and differed by location, with associations being primarily in the left tails in 
panel d, and primarily in the right tails in panel e (SI section \ref{SI-SIsect:planktonstats}). 
One can see these asymmetries by noting that the open points in panel d, which are in the left halves of the
distributions, are more tightly clustered around the line $y=x$ than are the closed points,
which are in the right halves of the distributions; and \emph{vice versa} for panel e.
(f-h) Lottery model population simulations% using $(B_1,B_2)$ 
obtained by transforming the respectively corresponding variables (a-c; see Introduction).
See Table 1 for used annotations and abbreviations.
\label{fig:pedagog1}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/theory_fig_col2.pdf}
\caption[Visual guide to removing asymmetries of association]{Illustration of how asymmetric tail associations are removed while retaining marginal distributions and correlation 
between variables (see Theory). (a) An example original bivariate distribution $(E_i,C_{i \bs i})$, with ATAs present. 
(b) Marginals are now normalized. The variables $E_i$ and $C_{i \bs i}$ are individually transformed to have standard normal 
marginals through a composition of the inverse of the standard normal cumulative distribution function (CDF), 
$\varphi^{-1}$, and 
each variable's CDF, $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively.  
(c) A newly generated bivariate normal distribution, $(e_i,c_{i \bs i})$, with correlation $\ddot{\rho}=\rho$
selected to match that of the original distribution (see Theory). (d) Two new variables, 
$E_i^{\nat}$ and  $C_{i \bs i}^{\nat}$,
which have the same marginals and correlation as the original $E_i$ and $C_{i \bs i}$ variables of (a) but 
which have symmetric association structure. These are the distributions used to compute ATA effects, 
$\Delta_i^{[EC]}$ (Theory). 
These variables are obtained by transforming the marginals from (c), 
$e_i$ and $c_{i \bs i}$, using the inverses of the functions used in (b).
Panel (e) shows the distribution $(E_i^\shp,C_{i \bs i}^\shp)$, defined in Theory
following \cite{Ellner_2016}, for comparison. These are the distributions used to compute storage effects,
$\Delta_i^{(EC)}$ (Theory).
For illustration purposes we started with beta distributed $E_i$ and $C_{i \bs i}$. Contour lines are 
$\log_{10}$ of probability density on all panels. 
See Table 1 for used annotations and abbreviations.
\label{fig:pedagog2}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLLN.pdf}
\caption{The contribution of ATAs to species coexistence was sometimes, but not always, 
meaningfully large compared to 
other components of the MCT decomposition for the log-normal fecundities lottery model.
All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel, 
$\delta$ and $\mu_1-\mu_2$ 
are fixed at values specified in the margins, and GWR and components of the decomposition
are plotted against $\sigma$. The contribution of ATAs ($\Delta_i^{[EC]}$) sometimes, but not always, 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the GWR line falls 
between the ATA line and the $x$ axis, indicating that, for those parameter values, ATAs 
make the difference between species
coexistence and exclusion of the weaker competitor. These instances, which are shaded, correspond to
ATA rescue or exclusion (Theory). Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3,5)}, so are not displayed. This plot uses $q_{ij}=1$. An 
analogous plot (SI Fig. \ref{SI-fig:fig3_alt}) uses 
$q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$; though results differed
in detail, main conclusions were the same. 
See Table 1 for used annotations and abbreviations. \label{fig:fig3}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLB.pdf}
\caption{The contribution of ATAs to species coexistence was often meaningfully large compared to 
other components of the MCT decomposition for the beta fecundities lottery model.
All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel,
$\delta$ is fixed at the value specified in the upper margin, 
and GWR and components of the decomposition
are plotted against the upper bound ratio, $\eta_2/\eta_1$. The contribution of ATAs ($\Delta_i^{[EC]}$) frequently 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the GWR line falls 
between the ATA line and the $x$ axis, indicating that, for those parameter values, ATAs make 
the difference between species
coexistence and exclusion of the weaker competitor. These instances, which are shaded, correspond to
ATA rescue or exclusion (Theory). Standard errors of plotted quantities 
were never greater than \Sexpr{round(se4,5)}, so are not displayed. 
This plot uses $q_{ij}=1$. 
An analogous plot (SI Fig. \ref{SI-fig:fig4_alt}) uses 
$q_{ij}=\eta_i/(\eta_j(1-\delta)+\eta_i \delta)$; 
though results differed in detail, main conclusions were the same.
See Table 1 for used annotations and abbreviations. }\label{fig:fig4}
\end{figure}

@@

\noindent Figure 5 caption. ATA effects and ATA rescue in the diatom system. 
The contribution of ATAs to the GWR of \emph{F. crotonensis}
in the diatom system of \cite{Descamps_2005} was often meaningfully large compared to 
other components of the MCT decomposition (a-c). Panels a-c show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). 
For each panel, 
two of the three parameters defining the temperature oscillation (the amplitude, $a$; period, $P$; 
and mean, $\theta_0$; see Methods) were fixed at the values used by 
\cite{Descamps_2005} ($a=6^\circ$C, $P=60$d, $\theta_0=18^\circ$C), and 
GWR and components of the MCT decomposition
were plotted against the third parameter. The original values of the varying parameters 
are indicated by vertical dotted lines. 
ATA contributions ($\Delta_i^{[EC]}$)
had magnitude comparable to other components, 
indicating that ATAs can contribute meaningfully
to coexistence in this system. In some instances, the GWR line falls 
between the ATA line
and the horizontal axis, indicating that, for those parameter values, ATA rescue occurs. 
Standard errors of plotted quantities were always less then \Sexpr{round(se5,5)}, so are not shown.
Panels d-f show bivariate plots (Methods) for which the third parameter was 
fixed at the values used by \cite{Descamps_2005}. ATAs are again often an important factor helping
determine coexistence or exclusion, and ATA rescue happened for an appreciable fraction of the
explored parameter space.
For the parameters we considered, GWR of the other diatom species in the system, 
\emph{C. pseudostelligera}, was always positive both including and
excluding the effects of ATAs, so GWR for \emph{F. crotonensis}, decomposed here, is a 
sufficient coexistence metric. See Table 1 for used annotations and abbreviations.
